ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming

### 2011-12-11 papers/IJCSE/foropencl.tex
more edits from bob

The sheer complexity of programming for clusters of many or multi-core
processors with tens of millions threads of execution makes the simplicity of
the data-parallel model attractive.  The increasing complexity of
today's applications (especially in light of the increasing complexity
of the hardware) and the need for portability across architectures
make a higher-level and simpler programming model like data-parallel
attractive.

>>>>

The sheer complexity of programming for clusters of many or multi-core
processors with tens of millions threads of execution makes the simplicity of
the data-parallel model attractive.  The increasing complexity of
today's applications (especially in light of the increasing complexity
of the hardware) and the need for portability across architectures
make a higher-level and simpler programming model like the data-parallel
model attractive.

### 2011-12-11 papers/IJCSE/programming-model.tex
more edits from bob

It should be noted that the function {\tt region\_cpy} is pure and thus can be
called from within a pure kernel procedure, and {\tt
region\_ptr} is impure because it aliases the array parameter.  However as will
be shown below, the usage of {\tt region\_ptr} is constrained so that it does not
introduce side effects in the functions that call it.  These two functions are
part of the language recognized by the compiler and though {\tt region\_cpy}
returns a copy of a portion of an array \emph{semantically}, the compiler is not
forced to actually make a copy and is free to enforce copy semantics through
other means.  In addition to these two new functions, ForOpenCL provides the
compiler directive, {\tt \!\$OFP PURE, KERNEL}, which specifies that a pure subroutine
can be transformed to an OpenCL kernel and that the subroutine is pure
except for calls to {\tt region\_ptr}.  These directives are not strictly
necessary for the technique described in this paper, but aid in automated
identification of specific kernels to be transformed to OpenCL.  A directive-free
implementation would require the transformation tool be provided the set of 
kernels to work via a user defined list.

>>>>

It should be noted that the function {\tt region\_cpy} is pure and
thus can be called from within a pure kernel procedure, and {\tt
  region\_ptr} is impure because it aliases the array parameter.
However as will be shown below, the usage of {\tt region\_ptr} is
constrained so that it does not introduce side effects in the
functions that call it.  These two functions are part of the language
recognized by the compiler and though {\tt region\_cpy} returns a copy
of a portion of an array \emph{semantically}, the compiler is not
forced to actually make a copy and is free to enforce copy semantics
through other means within the pure constraint.  In addition to these
two new functions, ForOpenCL provides the compiler directive, {\tt
  \!\$OFP PURE, KERNEL}, which specifies that a pure subroutine can be
transformed to an OpenCL kernel and that the subroutine is pure except
for calls to {\tt region\_ptr}.  These directives are not strictly
necessary for the technique described in this paper, but aid in
automated identification of specific kernels to be transformed to
OpenCL.  A directive-free implementation would require the
transformation tool be provided the set of kernels to work via a user
defined list.

### 2011-12-11 papers/IJCSE/shallow-water.tex
more edits from bob

\noindent
where h = height of water column (mass), $u$ = x velocity, $v$ =
y velocity, and $g$ = gravity. The height $h$ can be used for mass
because of the simplification of a unit cell size and a uniform water
density. Another simplifying assumption is that the water depth is
small in comparison to length and width and so velocities in the z-direction
can be ignored. A fixed time step is used for simplicity though it
must be less than $dt \leqq dx / (\sqrt{gh}+|u|)$ to fulfill the CFL
condition.

The numerical method is a two-step Lax-Wendroff scheme. The method
has some numerical oscillations with sharp gradients but is adequate
for simulating smooth shallow-water flows. In the following explanation,
$U$ is the conserved state variable at the center of the cell. This
state variable, $U$ $=(h,hu,hv)$ in the first term in the equations
below. $F$ is the flux quantity that crosses the boundary of the cell
and is subtracted from one cell and added to the other. The remaining
terms after the first term are the flux terms in the equations above
with one term for the flux in the x-direction and the next term for
the flux in the y-direction. The first step estimates the values a
half-step advanced in time and space on each face, using loops on
the faces.\begin{eqnarray*}
U_{i+\frac{1}{2},j}^{n+\frac{1}{2}} & = & (U_{i+1,j}^{n}+U_{i,j}^{n})/2+\frac{\triangle t}{2\triangle x}\left(F_{i+1,j}^{n}-F_{i,j}^{n}\right)\\
U_{i,j+\frac{1}{2}}^{n+\frac{1}{2}} & = & (U_{i,j+1}^{n}+U_{i,j}^{n})/2+\frac{\triangle t}{2\triangle y}\left(F_{i,j+1}^{n}-F_{i,j}^{n}\right)\end{eqnarray*}

>>>>

\noindent
where h = height of water column (mass), $u$ = x velocity, $v$ =
y velocity, and $g$ = gravity. The height $h$ can be used for mass
because of the simplification of a unit cell size and a uniform water
density. Another simplifying assumption is that the water depth is
small in comparison to length and width and so velocities in the z-direction
can be ignored. A fixed time step is used for simplicity though it
must be less than

\begin{eqnarray*}
dt \leqq \frac{dx}{\left(\left(\sqrt{gh}+|u|\right) + \left(\sqrt{gh}+|v|\right)\right)}
\end{eqnarray*}

\noindent to fulfill the CFL condition.

The numerical method is a two-step Lax-Wendroff scheme. The method
has some numerical oscillations with sharp gradients but is adequate
for simulating smooth shallow-water flows. In the following explanation,
$U$ is the conserved state variable at the center of the cell. This
state variable, $U$ $=\{h,hu,hv\}$ in the first term in the equations
below. $F$ is the flux quantity that crosses the boundary of the cell
and is subtracted from one cell and added to the other. The remaining
terms after the first term are the flux terms in the equations above
with one term for the flux in the x-direction and the next term for
the flux in the y-direction. The first step estimates the values a
half-step advanced in time and space on each face, using loops on
the faces.\begin{eqnarray*}
U_{i+\frac{1}{2},j}^{n+\frac{1}{2}} & = & (U_{i+1,j}^{n}+U_{i,j}^{n})/2+\frac{\triangle t}{2\triangle x}\left(F_{i+1,j}^{n}-F_{i,j}^{n}\right)\\
U_{i,j+\frac{1}{2}}^{n+\frac{1}{2}} & = & (U_{i,j+1}^{n}+U_{i,j}^{n})/2+\frac{\triangle t}{2\triangle y}\left(F_{i,j+1}^{n}-F_{i,j}^{n}\right)\end{eqnarray*}
